---
title: "Monetary policy and asset pricing - Assignment no. 4"
author: "Federico Vicentini"
date: "06/11/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pandoc)
```



```{r p0, message=FALSE, echo=FALSE}
# Clear the variables
rm(list = ls())

# Install packages
packages <- c("matlib", "rmarkdown", "tinytex","plotly","rgl","knitr","pandoc")
new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages)
invisible(lapply(packages, library, character.only = TRUE))
# Load packages
library(matlib)
library(scatterplot3d)
library(plotly)
library(rgl)
library(knitr)
options(digits = 15) 
library(pandoc)
```


## Exercise 1:

```{r p1, message=FALSE, echo=TRUE}
# Clear the variables
rm(list = ls())


# Set the working directory to source file location with
# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
setwd(getwd())

# ~  (this character will be needed later for copypasting)
phi <- 0.3
matrix <- matrix(0,2,2, dimnames=list(c("h","l"),c("h","l")))
matrix[1,] <- c(phi, (1-phi))
matrix[2,] <- c((1-phi), phi)
kable(matrix,   format = "pipe",
                row.names=TRUE, 
                col.names=c("h","l"))

beta <- 0.96
gamma <- 2
h <- 1.05
l <- 0.97
```

Now the exogenous endowment shock
is a two states Markov process, defined as:

$$\frac{\Omega_{t}}{\Omega_{t-1}}=\omega_{t}\quad 
  with \; \omega_{t}\epsilon (h,l)$$

With the following symmetric transition matrix:

$$\Pi = \left[\begin{matrix}\phi & (1-\phi) \\ 
                            (1-\phi) & \phi \\ \end{matrix}\right]$$

Let $\beta = 0.96$, $\gamma=2$, $h=1.05$, $l=0.97$ and $\phi=0.3$.

Knowing this, we cannot use the same 4-cases framework 
we've seen in class, as the probability of finding ourselves
in state h or state l now depends on the state of the economy in the
previous period.

Thus, we will have two different yield curves, conditional on the state
of the Markov process we are starting with at time t. Let's then compute
the yield curve for each of the two cases.

## CASE 1: (t = state h)

In this case, we start in h at time t, so the probability
matrix (at times $t+1$ and $t+2$) will look like this:

$$\left[\begin{matrix}\phi      & (\phi*\phi)+(1-\phi)*(1-\phi) \\ 
                      (1-\phi)  & \phi*(1-\phi)+(1-\phi)*\phi\\ 
        \end{matrix}\right]$$

Looking at the first column it's obvious that, since we find ourselves 
in $h$ at time $t$, in time $t+1$ we will be still in $h$ with probability
$\phi$ and we will switch to $l$ with probability $(1-\phi)$. 
Looking at the second column, we have to think about the ways through
which we could find ourselves in h at time $t+2$: either we were in $h$
at time $t$ and we remained there (probability equal to $(\phi*\phi)$) or 
we were in $l$ at time $t$ and we switched states (probability equal to 
$(1-\phi)*(1-phi)$). Summing the two probabilities we find the total probability
of finding ourselves in state $h$ at time $t+2$. The same exact
reasoning applies to the probability of finding ourselves in state $l$.


We can rewrite it as:


$$\left[\begin{matrix}\phi      & 2\phi^{2}+1-2\phi\\ 
                      (1-\phi)  & 2\phi(1-\phi)\\ 
        \end{matrix}\right]$$

Then, we compute the yield at time $t+1$ and $t+2$ using the following
code and the two equations:

$$y_{t+1}=-ln \left(\beta*\left(p_{1}(h)*(h)^{-\gamma} +
                                p_{1}(l)*(l)^{-\gamma}
              \right)\right)$$
$$y_{t+2}=-\frac{1}{2}ln \left(\beta^{2} 
                      \left(p_{1}(h) p_{2}(h) (h^{2})^{-\gamma} +
                            p_{1}(l) p_{2}(l) (l^{2})^{-\gamma} +
                            p_{1}(h) p_{2}(l) (hl)^{-\gamma} +
                            p_{1}(l) p_{2}(h) (lh)^{-\gamma}
              \right)\right)$$

We wanted also to have a measure of the short-term interest rate in 
period t. In the real world, this is the overnight interest rate, set
by the Central Bank. To make an approximation, we used the same equation
used for $y_{t+1}$ setting $p_{1}(h)=1$ since we know we are in $h$ at
time $t$. Also, $\beta$ has exponent $0$ so it is equal to $1$.
Thus, equation for our "short-term" interest rate is:

$$y_{t}=-ln \left((h)^{-\gamma}\right)$$

Let's look now at the code and the output:

```{r p2, message=FALSE, echo=TRUE, dev="svg"}

#CASE 1: 
#In this case we find ourselves in h at time t

pi <- matrix(NA, 2,2)
pi[1,]=c(phi, 2*(phi)^2+1-2*phi)
pi[2,]=c(1-phi, (2*phi)-2*(phi)^2)

yt_1=-log(beta*(pi[1,1]*(h)^(-gamma)+pi[2,1]*(l)^(-gamma)))
yt_2=-0.5*log(((beta)^2)*(pi[1,1]*pi[1,2]*(h*h)^(-gamma) + 
                          pi[2,1]*pi[2,2]*(l*l)^(-gamma) + 
                          pi[1,1]*pi[2,2]*(h*l)^(-gamma) + 
                          pi[2,1]*pi[1,2]*(l*h)^(-gamma)))

yieldcurve=c(yt_1,yt_2)
plot(yieldcurve, type="o",  main="Yield curve with t = h",
                            xlab= "Time",
                            ylab= "Yield",
                            col="dark green")


yt_0=-log((h)^(-gamma))
yieldcurve=c(yt_0,yt_1,yt_2)
plot(yieldcurve, type="o",  main="Yield curve with t = h",
                            xlab= "Time",
                            ylab= "Yield",
                            col="dark green")

yieldcurve * 100

```

In this case, the yield curve starts at about $9.75\%$, then it presents
a downward slope going to $2.49\%$ in the second period (indicating the
high probability of a recession in $t+1$) and then going up again to a yield
of $4.68\%$ in $t+2$ (indicating the prediction of a recovery in the second
period).

We observe that even if the curve predicts a recovery in the long term, does
not return to the first-period yield level. We believe this is because 
in the starting period the state $h$ is certain, while in the second it is 
only slightly more likely than a recession (looking at our results, we have $58\%$
probability of being in the high state), thus the yield is lower.



## CASE 2: (t = state l)

In the second case the procedure remains exactly identical, but the
two rows of the probability matrix are inverted with respect to the first case.

So, we report below the code and the related plots:

```{r p3, message=FALSE, echo=TRUE, dev="svg"}

#CASE 2: 
#In this case we find ourselves in l at time t

pi2 <- matrix(NA, 2,2)
pi2[1,]=c(1-phi, (2*phi)-2*(phi)^2)
pi2[2,]=c(phi, 2*(phi)^2+1-2*phi)

yt_1l=-log(beta*(pi2[1,1]*(h)^(-gamma)+pi2[2,1]*(l)^(-gamma)))
yt_2l=-0.5*log(((beta)^2)*(pi2[1,1]*pi2[1,2]*(h*h)^(-gamma) + 
                          pi2[2,1]*pi2[2,2]*(l*l)^(-gamma) + 
                          pi2[1,1]*pi2[2,2]*(h*l)^(-gamma) + 
                          pi2[2,1]*pi2[1,2]*(l*h)^(-gamma)))

yieldcurve_l=c(yt_1l,yt_2l)
plot(yieldcurve_l, type="o",  main="Yield curve with t = l",
                              xlab= "Time",
                              ylab= "Yield",
                              col="red")


yt_0l=-log((l)^(-gamma))
yieldcurve_l=c(yt_0l,yt_1l,yt_2l)
plot(yieldcurve_l, type="o",  main= "Yield curve with t = l",
                              xlab= "Time",
                              ylab= "Yield",
                              col="red")


yieldcurve_l * 100
```


In this case, the yield curve starts at about $-6.09\%$, then it presents
an upward slope going to $8.88\%$ in the second period (indicating the
high probability of an expansion in $t+1$) and then going down again to a yield
of $6.58\%$ in $t+2$ (indicating the prediction of a recession in the second
period).

We observe that even if the curve predicts a recession in the long term, it does
not return to the first-period yield level. We believe this is because 
in the starting period the state $l$ is certain, while in the second it is 
only slightly more likely than an expansion (looking at our results, we have $58\%$
probability of being in the low state), thus the yield is naturally higher.

Note that in both cases the yield $y_{t}$ is practically a "fictional" one. In the first 
case, the yield is much higher than what we would typically see in the real world economy, since
we do not ever have the complete certainty of being in a high state of growth, thus the real yield is
going to be much lower. In the second case, the yield is much lower than what we would observe. 
Also, let's recall that in reality, these short-term interest rate are mandated by the CBs, not by
the market itself.


## Comment on both cases

We can observe that, even if the the only difference in the probability matrix
in the two cases are the inverted rows, the yields are not "inverted".
In fact, the yield curve of the first case is always below the one for the 
second case (excluding $y_{t}$).

```{r p4, message=FALSE, echo=FALSE, dev="svg"}
plot(yieldcurve,  main= "Graph of both cases",
                  xlab= "Time",
                  ylab= "Yield",
                  col= "dark green",
                  type="o")
par(new=TRUE)
plot(yieldcurve_l,  col="red", 
                    type="o", 
                    main ="", 
                    xlab="", 
                    ylab="", 
                    axes=FALSE)
```

In fact, we cannot even state that the slope of the curve is the same in absolute value.
The two yield curves present different slopes, and we can motivate this by considering the fact
that, since we are dealing with a two-period framework, losing $3\%$ and then gaining $5\%$ (which
is the equivalent of ending up in $l$ and then going to $h$) does not result in the same
gains as the opposite path (gaining $5\%$ and then losing $3\%$). If we add to this the fact that
gains and losses are not equal in magnitude, we have a good explanation of why the two curves
behave in a different way.

# EXERCISE 2

```{r ex2, include=TRUE}
```